Applying State-free Reversible VAMPNets and Markov State Models to Learn Dynamics of DNA Oligonucleotides

### 2020-12-18 manuscript.tex
Updates from Overleaf

Despite rapid advances in DNA nanotechnology and a robust understanding of the associated thermodynamics, the sequence-dependent mechanisms of DNA hybridization are not fully understood. In this work, we investigate these dynamics by performing equilibrium coarse-grained simulations of oligonucleotide sequences with varied G:C placement. We employ State-Free Reversible VAMPnets to directly learn the slowest dynamical modes of each sequence and optimize Markov State Model (MSM) construction. Furthermore, we perform elevated temperature simulations to recapitulate temperature-jump IR and FTIR data collected on the oligonucleotides. For repetitive sequences, we find a spectrum of slow dynamics associated with out-of-register base pairing and kinetically relevant transitions between these states. In contrast, G:C pairs near the center of the duplex induce more pronounced fraying dynamics through which hybridization and dissociation are facilitated. In both cases, these mechanisms deviate from an “all-or-nothing” hybridization model. Our computational predictions show agreement with experiments, and provide new fundamental understanding of the sequence-dependent kinetics and mechanisms of DNA hybridization.

@@

Over the last couple decades, DNA has proved to be much more than a vessel for genetic information. From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications.\citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine} In particular, DNA dynamical behavior has become increasingly important in developing these technologies and understanding fundamental biological processes such as transcription and gene regulation.\citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment} A thorough understanding of the hybridization process -- the formation of a DNA duplex from two single strands -- and its associated dynamics is integral to further advancements in these fields. The effect of sequence on hybridization thermodynamics has been rigorously explored, and nearest-neighbor calculations can account for mismatched pairs, dangling ends, and other non-native bonding effects.\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications.\citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology} Although many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood.\citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon} Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if long-lived metastable states facilitate the transition.\citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}

Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly perturb DNA and monitor relaxation to a new equilibrium.\citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization} More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers.\citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution} Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail.\citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques.\citep{Phys2014} Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling.\citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative.\citep{Araque2016LatticeCooperativity, Phys2019} Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} Taken together, most experimental and computational work have studied the hybridization process as a relaxation to a chosen equilibrium or in a biased fashion.

Recent studies have coupled experimental techniques with machine learning (ML) and  MD simulations to investigate and predict sequence-dependent kinetics.\citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence} Where these studies focus on association and dissociation kinetics alone, we seek to broaden our analysis into higher order dynamical processes and metastable intermediates. For example, stable out-of-register or "shifted" base pairing in repetitive sequences have been documented in previous computational studies.\citep{Phys2014, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} The extent to which these states are kinetically relevant is difficult to determine experimentally, and appears highly sequence-dependent. Frayed structures and dynamics have also been investigated in numerous computational and experimental studies.\citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} Sanstead et al. highlighted the role of these dynamics during duplex dissociation, where the stability and relaxation timescale of frayed states was dictated by G:C base placement 10-mer oligonucleotides.\citep{Sanstead2016} Together, these dynamics represent substantial deviations from all-or-nothing behavior that can be variably expressed via subtle sequence differences in short oligomers.

In this work, we study the same four sequences explored by Sanstead et al. in an effort to uncover sequence-dependent dynamics and their relation to metastable structures mentioned above.\citep{Sanstead2016} We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to perform unbiased simulations of hybridization and dissociation behavior near each sequence's melting temperature.\citep{Hinckley2013AnHybridization} We synthesize the simulation trajectories using Markov State Models (MSMs) to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics.\citep{Pande2010EverythingAsk} MSMs have recently been implemented to study mechanisms of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies.\citep{Jin2019, Xiao2019} \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA oligonucleotides and found that stacking timescales are highly sequence dependent. We take a similar approach to study 10-mer DNA oligonucleotides and implement state free-reversible VAMPnets (SRVs) within the MSM parameterization pipeline to construct higher resolution kinetic models than are achievable using standard time-lagged independent components analysis (tICA) based approaches.

@@

We initialized four self-complementary double-stranded sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- in a periodic box of dimensions 7.8 nm.\citep{Sanstead2016} We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff. We ran our simulations in the NVT ensemble. Temperature was controlled by a Langevin thermostat that models interactions with an implicit water solvent.\citep{Schneider1978Molecular-dynamicsTransitions} Simulations were conducted at empirically determined  melting temperatures in order to observe frequent transitions between dissociated and hybridized states. We used a 20 fs time step and $1.3\times10^{9}$ steps in each simulation, saving frames every 100 ps. For each sequence, 40 simulations were performed in parallel, consuming about 24 serial CPU-hours on 28xIntel E5-2680v4 processors per simulation. Half of runs were initialized from the native hybridized state and half from a dissociated state. The first 10000 frames (1 $\mu$s of simulation time) of each simulation were removed, resulting in 40 x 250000 frames and a total of 1 ms simulation time per sequence. We observed between 55-100 hybridization/dissociation events for each of the four sequences.

@@

We followed the SRV-MSM analysis pipeline described above to generate sequence-dependent kinetic models for each sequence. We identified seven kinetically relevant states that are captured within the resolution of the model. Figure \ref{fig:allseq_table} shows cartoon and molecular rendering for each of these states as well as their sequence-specific stationary probabilities. By row, these include the fully hybridized state (H) in which all native base pairings are intact, four "shifted" states (5S2, 3S2, 5S4, 3S4) where complementary base pairs form out-of-register, a frayed state (F4), unique to GC-core, in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D). Figure \ref{fig:allseq_transition} shows how these states fit into sequence-specific low-dimensional representations and kinetic models. Arrows between states show the probability of a transition to another state within the MSM lag time. We now proceed to analyse the sequence-dependent thermodynamics and kinetic features of each MSM.

@@

Nearest neighbor (NN) models have been experimentally optimized to account for stacking contributions of native pairs, internal mismatches, and dangling ends.\citep{Santalucia2004TM} The effects of inert tails -- free bases that extend beyond the dangling end -- have been shown to destabilize the duplex but are not included in NN models.\citep{Michele2014EHybridization} It is informative to compare our MSM predictions against NN calculations to understand how inert tail and kinetic effects may cause deviation from these predictions. In particular, we examine  the out-of-register AT-all and GC-end states which are amenable to NN comparisons (figure \ref{fig:NN_table}). For AT-all, NN calculations  predict that conformations in the 5' shifted states (5S2 and 5S4) are more energetically favorable than those in the 3' shifted state (3S2 and 3S4). We find qualitative agreement to our macrostate free energies, where inert tail effects likely contribute a 4-6.5 kJ/mol increase in free energy compared to dangling end predictions alone. For GC-end, we consider C:T and G:A mismatches as non-interacting dangling ends such that each shifted conformation has four total dangling ends. Based on this treatment, NN calculations yield higher overall free energies due to fewer native base pair contacts.  Contrary to our simulation results, however, the NN model predicts the 3S2 state to be more stable than 5S2 state. Moreover, we do not observe a stable 3S4 cluster for GC-end. This indicates that 5' vs. 3' inert tail differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN stacking effects alone.\citep{Doktycz1990ThermodynamicATGC}

@@

% This leads us to believe that 5' vs. 3' differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN effects \citep{Doktycz1990ThermodynamicATGC}.

@@

In addition to state probabilities and free energy approximations, the coarse-grained MSM yields valuable discrete kinetic information in the form of transition probabilities between states. For AT-all, we observe approximately equal probability of transitioning from D to any other state. Once a transition has been made, however, the 5' vs. 3' overhang and degree of shifting play an important role in determining whether the duplex will continue to shift out-of-register or re-dissociate. Transition probabilities are higher when moving towards a more aligned state than towards a more shifted state -- 5S2 $\rightarrow$ H is more favorable than 5S2 $\rightarrow$ 5S4 -- suggesting that more shifted states play a greater role in facilitating the hybridization process than dissociation. Furthermore, we see equal or higher transition probability from shifted states to the dissociated state (5S2 $\rightarrow$ D) than to more aligned states (5S2 $\rightarrow$ H), indicating that the shifting-hybridization process is frequently disrupted by complete dissociation. In particular, we observe that the transition probability 3S4 $\rightarrow$ D is 10x higher than the 3S4 $\rightarrow$ 3S2 path towards native hybridization.

@@

Experimentally, out-of-register states are suspected to occur, but are difficult to capture due to short lifetimes and similarity in response to other processes such as fraying. To explicitly minimize out-of-register base pairing, similar AT repeat motif sequences have been padded by GCG clamps during experimental analysis.\cite{Wyer2014KineticsAT-tracts} Recent all-atom results identified analogous out-of-register states as "deep kinetic traps" along the hybridization pathway for repetitive dGCGCGC hexamers.\citep{Xiao2019} Contrary to our kinetic results, however, "slithering" mechanisms along out-of-register tracts were not observed to be a dominant pathway, especially when compared to the high rates of slithering exhibited by the homogenous dGGGGGG strand. It is unclear whether these differences were a consequence of varied simulation conditions or whether GC repeat motifs are less susceptible to direct out-of-register transitions compared to AT motifs. This is possible given that stronger hydrogen binding in GC motifs \citep{Yakovchuk2006Base-stackingHelix,Zacharias2020Base-PairingFormation} may prevent fluctuation-driven rearrangement, however further computational and experimental studies are required to verify these differences.

@@

The GC-core sequence represents a departure from the dominant shifting dynamics observed for AT-all and GC-end. Instead, the dynamical analysis describes a hybridization/dissociation pathway facilitated by a unique, highly frayed state (F4). Previous studies suggests that once key native contacts are made, the zippering mechanism ensures that the helix will quickly form outward.\citep{Romano2013DNADependence, Yin2011KineticsHybridization} Our results indicate, however, that the relative instability of A:T bonds compared the GC core can interrupt this process and form a longer lived metastable state. This occurs during the dissociation process as well, where one half of the A:T base contacts are entirely broken for a substantial period of time before the full dissociation event occurs. We observe these events to occur with equal probability on either permutable end of the helix.

@@

Given the lack of a repetitive AT interior (as in AT-all and GC-end) or exterior (as in GC-core), we expect more canonical dynamics from GC-mix. Although GC-mix dynamics are most similar to those of GC-core, we did not observe a converged slow mode corresponding to multi-base fraying behavior for GC-mix. Instead, we observed two modes converge, corresponding to the association/dissociation dynamic and diffusive behavior while strands are dissociated, respectively. Given that this second mode did not inform the hybridization process, we designated this transitions as effectively two-state within the resolution of our model. We did, however, observe substantial fraying of the two AT termini in the simulation data. Although these frayed states may be too short-lived to resolve a distinct slow mode, this behavior shows qualitative agreement with experimental analysis of this sequence which attributed fraying prior to dissociation as a deviation from all-or-nothing behavior.\citep{Sanstead2016}  While AT-termini fraying is surely a prerequisite to dissociation, we find these states to be so common and fleeting that very few progress to full dissociation compared to the F4 state. Furthermore, one or two base pair fraying does not fundamentally disrupt the helix in such a way that its re-formation is kinetically inhibited by the intermediate structures we present for GC-core.

@@

Although we were able to obtain improved resolutions on several relevant dynamics by using a shared lag time across sequences, we found it difficult to converge faster processes such as duplex nucleation and zippering. These processes are crucial to duplex formation, however they do not appear to be kinetically metastable or slow relative to other timescales of interest. Furthermore, they can initiate at various points along the strand, which, under out present featurization method, may appear as a collection of modes instead of as one distinct process. We also found the need to strike a balance between adequate sampling of hybridization events and frame save rate in order to maintain tractable SRV-MSM calculations. Indeed, we collected over 10 GB of equilibrium trajectory data for each sequence, and were working near memory limits when training SRVs and building SRV-MSMs.

>>>>

Despite rapid advances in DNA nanotechnology and a robust understanding of the associated thermodynamics, the sequence-dependent mechanisms of DNA hybridization are not fully understood. In this work, we investigate these dynamics by performing equilibrium coarse-grained simulations of oligonucleotide sequences with varied G:C placement. We employ State-Free Reversible VAMPnets to directly learn the slowest dynamical modes of each sequence and optimize Markov State Model (MSM) construction. Furthermore, we perform elevated temperature simulations to recapitulate temperature-jump IR and FTIR data collected on the oligonucleotides. For repetitive sequences, we find a spectrum of slow dynamics associated with out-of-register base pairing and kinetically relevant transitions between these states. In contrast, G:C pairs near the center of the duplex induce more pronounced fraying dynamics through which hybridization and dissociation are facilitated. In both cases, these mechanisms deviate from an “all-or-nothing” hybridization model. Our computational predictions show agreement with experiments and provide new fundamental understanding of the sequence-dependent kinetics and mechanisms of DNA hybridization.

@@

Over the last of couple decades, DNA has proven to be much more than a vessel for genetic information. From sensing to computing to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications.\citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine} In particular, DNA dynamical behavior has become increasingly important in developing these technologies and understanding fundamental biological processes such as transcription and gene regulation.\citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment} A thorough understanding of the hybridization process -- the formation of a DNA duplex from two single strands -- and its associated dynamics is integral to further advancements in these fields. The effect of sequence on hybridization thermodynamics has been rigorously explored, and nearest-neighbor calculations can account for mismatched pairs, dangling ends, and other non-native bonding effects.\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications.\citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology} Although many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood.\citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon} Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if long-lived metastable states facilitate the transition.\citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}

Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly perturb DNA and monitor relaxation to a new equilibrium.\citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization} More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers.\citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution} Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail.\citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques.\citep{Phys2014} Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling.\citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative.\citep{Araque2016LatticeCooperativity, Phys2019} Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} Taken together, most experimental and computational works have studied the hybridization process as a relaxation to a chosen equilibrium or in a biased fashion.

Recent studies have coupled experimental techniques with machine learning (ML) and  MD simulations to investigate and predict sequence-dependent kinetics.\citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence} Where these studies focus on association and dissociation kinetics alone, we seek to broaden our analysis into higher-order dynamical processes and metastable intermediates. For example, stable out-of-register or "shifted" base pairing in repetitive sequences have been documented in previous computational studies.\citep{Phys2014, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} The extent to which these states are kinetically relevant is difficult to determine experimentally, and appears highly sequence-dependent. Frayed structures and dynamics have also been investigated in numerous computational and experimental studies.\citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} Sanstead et al. highlighted the role of these dynamics during duplex dissociation, where the stability and relaxation timescale of frayed states was dictated by G:C base placement 10-mer oligonucleotides.\citep{Sanstead2016} Together, these dynamics represent substantial deviations from all-or-nothing behavior that can be variably expressed via subtle sequence differences in short oligomers.

In this work, we study the same four sequences explored by Sanstead et al. in an effort to uncover sequence-dependent dynamics and how they relate to metastable structures mentioned above.\citep{Sanstead2016} We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to perform unbiased simulations of hybridization and dissociation behavior near each sequence's melting temperature.\citep{Hinckley2013AnHybridization} We synthesize the simulation trajectories using Markov State Models (MSMs) to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics.\citep{Pande2010EverythingAsk} MSMs have recently been implemented to study mechanisms of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies.\citep{Jin2019, Xiao2019} \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA oligonucleotides and found that stacking timescales are highly sequence dependent. We take a similar approach to study 10-mer DNA oligonucleotides and implement state free-reversible VAMPnets (SRVs) within the MSM parameterization pipeline to construct higher resolution kinetic models than are achievable using standard time-lagged independent components analysis (tICA) based approaches.

@@

We initialized four self-complementary double-stranded sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- in a periodic box of dimensions 7.8 nm.\citep{Sanstead2016} We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff and ran our simulations in the NVT ensemble. Temperature was controlled by a Langevin thermostat that models interactions with an implicit water solvent.\citep{Schneider1978Molecular-dynamicsTransitions} Simulations were conducted at empirically determined  melting temperatures in order to observe frequent transitions between dissociated and hybridized states. We used a 20 fs time step and $1.3\times10^{9}$ steps in each simulation, saving frames every 100 ps. For each sequence, 40 simulations were performed in parallel, consuming about 24 serial CPU-hours on 28xIntel E5-2680v4 processors per simulation. Half of runs were initialized from the native hybridized state and half from a dissociated state. The first 10000 frames (1 $\mu$s of simulation time) of each simulation were removed, resulting in 40 x 250000 frames and a total of 1 ms simulation time per sequence. We observed between 55-100 hybridization/dissociation events for each of the four sequences.

@@

We followed the SRV-MSM analysis pipeline described above to generate sequence-dependent kinetic models for each sequence. We identified seven kinetically relevant states that are captured within the resolution of the model. Figure \ref{fig:allseq_table} shows cartoon and molecular renderings for each of these states as well as their sequence-specific stationary probabilities. By row, these include the fully hybridized state (H) in which all native base pairings are intact, four "shifted" states (5S2, 3S2, 5S4, 3S4) where complementary base pairs form out-of-register, a frayed state (F4), unique to GC-core, in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D). Figure \ref{fig:allseq_transition} shows how these states fit into sequence-specific low-dimensional representations and kinetic models. Arrows between states show the probability of a transition to another state within the MSM lag time. We now proceed to analyze the sequence-dependent thermodynamics and kinetic features of each MSM.

@@

Nearest neighbor (NN) models have been experimentally optimized to account for stacking contributions of native pairs, internal mismatches, and dangling ends.\citep{Santalucia2004TM} The effects of inert tails -- free bases that extend beyond the dangling end -- have been shown to destabilize the duplex but are not included in NN models.\citep{Michele2014EHybridization} It is informative to compare our MSM predictions against NN calculations to understand how inert tail and kinetic effects may cause deviation from these predictions. In particular, we examine  the out-of-register AT-all and GC-end states which are amenable to NN comparisons (figure \ref{fig:NN_table}). For AT-all, NN calculations  predict that conformations in the 5' shifted states (5S2 and 5S4) are more energetically favorable than those in the 3' shifted state (3S2 and 3S4). We find qualitative agreement to our macrostate free energies, where inert tail effects likely contribute a 4-6.5 kJ/mol increase in free energy compared to dangling end predictions alone. For GC-end, we consider C:T and G:A mismatches as non-interacting dangling ends such that each shifted conformation has four total dangling ends. Based on this treatment, NN calculations yield higher overall free energies due to fewer native base pair contacts.  Contrary to our simulation results, however, the NN model predicts the 3S2 state to be more stable than 5S2 state. Moreover, we do not observe a stable 3S4 cluster for GC-end. This indicates that 5' vs. 3' inert tail differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weigh NN stacking effects alone.\citep{Doktycz1990ThermodynamicATGC}

@@

% This leads us to believe that 5' vs. 3' differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weigh NN effects \citep{Doktycz1990ThermodynamicATGC}.

@@

In addition to state probabilities and free energy approximations, the coarse-grained MSM yields valuable discrete kinetic information in the form of transition probabilities between states. For AT-all, we observe approximately equal probability of transitioning from D to any other state. Once a transition has been made, however, the 5' vs. 3' overhang and degree of shifting play an important role in determining whether the duplex will continue to shift out-of-register or re-dissociate. Transition probabilities are higher when moving towards a more aligned state than towards a more shifted state -- e.g. 5S2 $\rightarrow$ H is more likely than 5S2 $\rightarrow$ 5S4 -- suggesting that more shifted states play a greater role in facilitating the hybridization process than dissociation. Furthermore, we see equal or higher transition probability from shifted states to the dissociated state (5S2 $\rightarrow$ D) than to more aligned states (5S2 $\rightarrow$ H), indicating that the shifting-hybridization process is frequently disrupted by complete dissociation. In particular, we observe that the transition probability 3S4 $\rightarrow$ D is 10x higher than the 3S4 $\rightarrow$ 3S2 path towards native hybridization.

@@

Experimentally, out-of-register states are suspected to occur but are difficult to capture due to short lifetimes and similarity in response to other processes such as fraying. To explicitly minimize out-of-register base pairing, similar AT repeat motif sequences have been padded by GCG clamps during experimental analysis.\cite{Wyer2014KineticsAT-tracts} Recent all-atom results identified analogous out-of-register states as "deep kinetic traps" along the hybridization pathway for repetitive dGCGCGC hexamers.\citep{Xiao2019} Contrary to our kinetic results, however, "slithering" mechanisms along out-of-register tracts were not observed to be a dominant pathway, especially when compared to the high rates of slithering exhibited by the homogenous dGGGGGG strand. It is unclear whether these differences were a consequence of varied simulation conditions or whether GC repeat motifs are less susceptible to direct out-of-register transitions compared to AT motifs. This is possible given that stronger hydrogen binding in GC motifs \citep{Yakovchuk2006Base-stackingHelix,Zacharias2020Base-PairingFormation} may prevent fluctuation-driven rearrangement, however further computational and experimental studies are required to verify these differences.

@@

The GC-core sequence represents a departure from the dominant shifting dynamics observed for AT-all and GC-end. Instead, the dynamical analysis describes a hybridization/dissociation pathway facilitated by a unique, highly frayed state (F4). Previous studies suggest that once key native contacts are made, the zippering mechanism ensures that the helix will quickly form outward.\citep{Romano2013DNADependence, Yin2011KineticsHybridization} Our results indicate, however, that the relative instability of A:T bonds compared the GC core can interrupt this process and form a longer lived metastable state. This occurs during the dissociation process as well, where one half of the A:T base contacts are entirely broken for a substantial period of time before the full dissociation event occurs. We observe these events to occur with equal probability on either permutable end of the helix.

@@

Given the lack of a repetitive AT interior (as in AT-all and GC-end) or exterior (as in GC-core), we expect more canonical dynamics from GC-mix. Although GC-mix dynamics are most similar to those of GC-core, we did not observe a converged slow mode corresponding to multi-base fraying behavior for GC-mix. Instead, we observed two modes converge, corresponding to the association/dissociation dynamic and diffusive behavior while strands are dissociated, respectively. Given that this second mode did not inform the hybridization process, we designated this transitions as effectively two-state within the resolution of our model. We did, however, observe substantial fraying of the two AT termini in the simulation data. Although these frayed states may be too short-lived to resolve a distinct slow mode, this behavior shows qualitative agreement with experimental analysis of this sequence which attributed fraying prior to dissociation as a deviation from all-or-nothing behavior.\citep{Sanstead2016}  While AT-termini fraying is surely a prerequisite to dissociation, we find these states to be so common and fleeting that very few progress to full dissociation compared to the F4 state. Furthermore, one or two-base-pair fraying does not fundamentally disrupt the helix in such a way that its re-formation is kinetically inhibited by the intermediate structures we present for GC-core.

@@

Although we were able to obtain improved resolutions on several relevant dynamics by using a shared lag time across sequences, we found it difficult to converge faster processes such as duplex nucleation and zippering. These processes are crucial to duplex formation, however they do not appear to be kinetically metastable or slow relative to other timescales of interest. Furthermore, they can initiate at various points along the strand, which, under out present featurization method, may appear as a collection of modes instead of as one distinct process. We also found the need to strike a balance between adequate sampling of hybridization events and frame save rate in order to maintain tractable SRV-MSM calculations. Indeed, we collected over 10 GB of equilibrium trajectory data for each sequence and were working near memory limits when training SRVs and building SRV-MSMs.
